Processing CDC hypertension and CDV Mortality disease data from adults in the USA exploratory analysis script

### 2025-05-02 code/analysis-code/statistical-analysis-v2-slim.qmd
Changes for final product

@@

```{r setup}
#load needed packages. make sure they are installed.
#install.packages("gt")
#install.packages("emmeans")
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidyverse)
library(tidymodels)
library(gt)
library(stringr)
library(patchwork)
library(emmeans)
library(sf) #for geospatial analysis
library(spdep)
library(spatialreg)
library(vip) # For variable importance plots
library(dplyr)
library(tidytext)
library(forcats)

@@

# Display and save the regression results
print(lm_results_year)
saveRDS(lm_results_year, file = here::here("products", "manuscript", "supplement", "lm_results_year.rds"))

@@

# Save as table
saveRDS(model_fit_lm, here::here("results", "tables", "model_fit_national.rds"))

@@

print(lm_plot_year)
ggsave(here::here("products", "manuscript", "supplement", "lm_plot_year.png"), plot = lm_plot_year)

@@

The linear regression model estimated a modest but statistically significant annual increase in national hypertension-related CVD mortality rates between 2000 and 2019 (slope = 1.40, p < 2.2e-16). However, as shown in the plot, the fitted trend line is visually flat due to the scale and large variation in county-level rates. The intercept value is not directly interpretable, but reflects the expected mortality rate at year zero in the model's coding. Despite the statistically significant trend, the magnitude of change is small relative to the observed range, suggesting that national mortality rates remained relatively stable over time. This model does not account for age or other demographic factors, motivating the more detailed modeling in the next section.

@@

# Display and save the GLM results
print(glm_results)
saveRDS(glm_results, file = here::here("products", "manuscript", "supplement", "glm_results.rds"))

@@

  print(glm_race_sex_plot)
  ggsave(here::here("products", "manuscript", "supplement", "glm_race_sex_plot.png"), plot = glm_race_sex_plot)

@@

The GLM including race/ethnicity and sex as predictors highlights large disparities in predicted mortality rates. Black populations had the highest predicted rates, significantly higher than all other groups (e.g., β = 205.8, p < 2e-16). Hispanic and Asian/Pacific Islander groups had lower mortality, though the Asian effect was not significant (p = 0.35). Males also showed significantly higher mortality than females (β = 20.3, p < 2e-16). Notably, the “Overall” category behaves inconsistently in plots due to it being modeled as a separate group rather than a true average, likely affected by unmodeled age variation. This supports the need to model interactions and stratify further by age.

@@

saveRDS(glm_results_interaction, file = here::here("products", "manuscript", "supplement", "glm_results_interaction.rds"))

# Plot the interaction model results
county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")
glm_interaction_plot <- ggplot(county_race_sex, aes(x = Year, y = predicted_mortalities_interaction, color = race_ethnicity)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "GLM Interaction Model: Mortality Trends by Year and Race", x = "Year", y = "Predicted Mortality Rate", color = "Race/Ethnicity") +
  scale_x_continuous(breaks = seq(min(county_race_sex$Year, na.rm = TRUE), max(county_race_sex$Year, na.rm = TRUE), by = 5)) +
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.key.size = unit(0.4, "cm"), legend.spacing.y = unit(0.2, "cm"), axis.text.x = element_text(angle = 45, hjust = 1))

print(glm_interaction_plot)
ggsave(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"), plot = glm_interaction_plot)

saveRDS(county_race_sex, file = here::here("results", "output", "county_race_sex_with_preds.rds"))
```

Including an interaction between year and race revealed significantly divergent temporal trends. Mortality rates among Black individuals showed the steepest annual decline (β = –7.06, p < 0.001), while American Indian and Alaska Native (AI/AN) populations exhibited the largest annual increase (β for year = 3.38, p < 0.001, with positive main effect and significant negative interactions for other groups). Asian and Pacific Islander, Hispanic, and White populations also showed declining trends, though to a lesser extent. These results highlight substantial heterogeneity in how hypertension-related CVD mortality changed over time across racial/ethnic groups.

@@

# Save results
saveRDS(glm_age_group_results, file = here::here("products", "manuscript", "supplement", "glm_age_group_results.rds"))

@@

# Save as table
saveRDS(model_fit_agegroup, here::here("results", "tables", "model_fit_agegroup.rds"))

@@

# Save plot
ggsave(here::here("products", "manuscript", "supplement", "glm_age_plot.png"), 
       plot = glm_age_plot, width = 9, height = 6)

@@

This model identifies age-specific trends.Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends.
Including an interaction between year and age group revealed that mortality rates are substantially higher in the 65+ age group across all years. While both age groups show increasing mortality over time, the slope is significantly steeper for older adults (interaction term β = 0.224, p < 0.001), suggesting accelerated increases in mortality among those aged 65 and older.

@@

# Save as manuscript-ready table
saveRDS(slope_table_age_fmt, here::here("results", "tables", "slope_table_agegroup_fmt.rds"))

# compare trends between groups (pairwise contrast)
slope_contrast <- contrast(em_trends_age, method = "pairwise")
summary(slope_contrast)

# Save
saveRDS(em_trends_age, file = here::here("products", "manuscript", "supplement", "emtrends_age_group.rds"))
saveRDS(slope_contrast, file = here::here("products", "manuscript", "supplement", "slope_contrast_age_group.rds"))

@@

# Save formatted contrast table
saveRDS(slope_contrast_age_fmt, here::here("results", "tables", "slope_contrast_agegroup_fmt.rds"))

@@

# Save figure to manuscript folder
ggsave(
  filename = here::here("results", "figures", "slope_plot_age_group.png"),
  plot = slope_plot_agegroup,
  width = 9, height = 6, dpi = 300)

# Save RDS for reproducibility
saveRDS(slope_plot_agegroup, file = here::here("results", "figures", "slope_plot_age_group.rds"))

@@

Estimated yearly changes in mortality rates were significantly higher for adults aged 65+ (slope = 1.26, 95% CI: 1.18–1.34) compared to those aged 35–64 (slope = 1.04, 95% CI: 0.96–1.12). The difference in slope between age groups was statistically significant (difference = -0.224, p < 0.001), indicating a steeper rise in mortality among older adults.

@@

print(spatial_plot_all)
ggsave(here::here("products", "manuscript", "supplement", "spatial_plot_all.png"), plot = spatial_plot_all)

@@

print(black_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "black_spatial_plot.png"), plot = black_spatial_plot)

@@

print(white_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "white_spatial_plot.png"), plot = white_spatial_plot)

@@

print(hispanic_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "hispanic_spatial_plot.png"), plot = hispanic_spatial_plot)

@@

print(native_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "native_spatial_plot.png"), plot = native_spatial_plot)

@@

print(asian_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "asian_spatial_plot.png"), plot = asian_spatial_plot)

@@

# Combine all of the spatial charts using patchwork
combined_spatial_plot <- (spatial_plot_all | black_spatial_plot | white_spatial_plot) /
  (hispanic_spatial_plot | native_spatial_plot |
  asian_spatial_plot)

# Print and save the combined spatial plot
print(combined_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"), 
       plot = combined_spatial_plot, width = 14, height = 10)

These plots are best viewed as images outside of R due to size.

@@

### Evaluating spatial clustering of Mortality Rates.
### 2.3: Global Moran’s I – All Races Combined

@@

# Save to results/tables
saveRDS(moran_table, here::here("results", "tables", "global_morans_i_summary.rds"))
saveRDS(moran_result, file = moran_path)

@@

The Global Moran’s I statistic (I = 0.0435, p < 0.001) indicates weak but statistically significant positive spatial autocorrelation in hypertension-related CVD mortality at the county level. There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S. counties — counties with high (or low) mortality tend to be near others with similarly high (or low) values. Although the magnitude is small (I = 0.0435), the pattern is considered statistically significant based on the p-value. This suggests that counties with high mortality rates tend to cluster geographically, rather than being randomly distributed. The presence of spatial autocorrelation supports further analysis using Local Moran’s I to identify specific high- and low-mortality clusters.

@@

# Save plot
ggsave(filename = here::here("products", "manuscript", "supplement", "lisa_significance_map_centroids.png"),
  plot = lisa_point_plot,
  width = 10,
  height = 6)

@@

ggsave(
  here::here("products", "manuscript", "supplement", "lisa_cluster_type_map.png"),
  plot = lisa_cluster_plot,
  width = 10, height = 6)

@@

#### Focusing at a State level Due to the size of the dataset it is challenging to work with the full dataset. 
For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step. Our previous analysis support that there is a general distribution of cases here, even if there are not distinct hotspots.

@@

ggsave(
  here::here("products", "manuscript", "supplement", "ga_lisa_cluster_plot.png"),
  plot = ga_lisa_plot,
  width = 8,
  height = 6)
```

@@

ggsave(filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
  plot = ms_lisa_plot,
  width = 8,
  height = 6)

@@

# Save plot
ggsave(
  filename = here::here("products", "manuscript", "supplement", "tx_lisa_cluster_plot.png"),
  plot = tx_lisa_plot,
  width = 9, height = 6)
```

@@

# Paths for saving spatial weights
ky_knn_path <- here::here("results", "output", "ky_knn_neighbors.rds")
ky_lw_path  <- here::here("results", "output", "ky_spatial_weights.rds")

@@

# Save
ggsave(
  here::here("products", "manuscript", "supplement", "lisa_cluster_kentucky.png"),
  plot = lisa_ky_plot, width = 9, height = 6)

@@

# Save the result
saveRDS(ky_moran_result, here::here("results", "output", "ky_global_morans_i.rds"))

@@

The Local Moran’s I analysis detected significant spatial clustering of mortality rates across U.S. counties. Nationally, over 40% of counties showed statistically significant clustering at p < 0.05. However, most of these represented spatial outliers—counties with high (or low) mortality rates surrounded by contrasting neighbors—rather than concentrated hotspots. Only a small number of “High–High” clusters (hotspots) were identified, indicating that extreme spatial concentrations of high mortality are rare.

@@

>>>>

This should be the third and final file to run when reproducing the full analysis. Thereafter you will get results in the manuscript/supplementary.

@@

```{r setup}
#load needed packages. make sure they are installed.
#install.packages("gt")
#install.packages("emmeans")
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidyverse)
library(tidymodels)
library(gt)
library(stringr)
library(patchwork)
library(emmeans)
library(sf) #for geospatial analysis
library(spdep)
library(spatialreg)
library(vip) # For variable importance plots
library(dplyr)
library(tidytext)
library(forcats)
library(webshot2)  # for gt image export
library(ggpubr)

@@

# Display and save the regression results
print(lm_results_year)
invisible(saveRDS(lm_results_year, file = here::here("results", "output", "statistical-analysis", "lm_results_year.rds")))

@@

# Save as table
invisible(saveRDS(model_fit_lm, here::here("results", "tables", "model_fit_national.rds")))

@@

print(lm_plot_year)
invisible(ggsave(here::here("results", "figures", "lm_plot_year.png"), plot = lm_plot_year))

@@

The linear regression model estimated a modest but statistically significant annual increase in national hypertension-related CVD mortality rates between 2000 and 2019 (slope = 1.40, p \< 2.2e-16). However, as shown in the plot, the fitted trend line is visually flat due to the scale and large variation in county-level rates. The intercept value is not directly interpretable, but reflects the expected mortality rate at year zero in the model's coding. Despite the statistically significant trend, the magnitude of change is small relative to the observed range, suggesting that national mortality rates remained relatively stable over time. This model does not account for age or other demographic factors, motivating the more detailed modeling in the next section.

@@

# Display and save the GLM results
print(glm_results)
invisible(saveRDS(glm_results, file = here::here("results", "output", "statistical-analysis", "glm_results.rds")))

@@

  print(glm_race_sex_plot)
  invisible(ggsave(here::here("results", "figures", "glm_race_sex_plot.png"), plot = glm_race_sex_plot))

@@

# Display and save the GLM results
print(glm_results)
invisible(saveRDS(glm_results, file = here::here("results", "output", "statistical-analysis", "glm_results.rds")))

# Add predicted values
county_race_sex$predicted_mortalities <- predict(glm_model, type = "response")

# Filter out "Overall" race_ethnicity values before plotting
plot_data <- county_race_sex %>%
  filter(race_ethnicity != "Overall")

# Create polished plot
glm_race_sex_plot <- ggplot(plot_data, aes(x = Year, y = predicted_mortalities, color = race_ethnicity)) +
  geom_line(size = 1.2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "GLM Predictions: Mortality Rates by Year and Race/Ethnicity",
    x = "Year",
    y = "Predicted Mortality Rate (per 100,000)",
    color = "Race/Ethnicity"
  ) +
  scale_x_continuous(
    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
  ) +
  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.2, "cm"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5))

# Display plot
print(glm_race_sex_plot)

# Save figure
ggsave(
  here::here("results", "figures", "glm_race_sex_plot.png"),
  plot = glm_race_sex_plot,
  width = 7.5,
  height = 5.5,
  dpi = 300)

# Fit GLM model (no interaction)
glm_model <- glm(mortalities ~ Year + race_ethnicity + sex, data = county_race_sex, family = gaussian())

# Save tidy model output
glm_results <- broom::tidy(glm_model)
print(glm_results)
invisible(saveRDS(glm_results, file = here::here("results", "output", "statistical-analysis", "glm_results.rds")))

# Predict mortality values from model
county_race_sex$predicted_mortalities <- predict(glm_model, type = "response")

# Remove "Overall" for plotting
filtered_data <- county_race_sex %>%
  filter(race_ethnicity != "Overall")

# Summarize observed means
observed_summary <- filtered_data %>%
  group_by(Year, race_ethnicity) %>%
  summarise(
    observed_mean = mean(mortalities, na.rm = TRUE),
    .groups = "drop")

# Summarize predicted means
predicted_summary <- filtered_data %>%
  group_by(Year, race_ethnicity) %>%
  summarise(
    predicted_mean = mean(predicted_mortalities, na.rm = TRUE),
    .groups = "drop")

# Join observed and predicted for plotting
plot_data <- left_join(observed_summary, predicted_summary, by = c("Year", "race_ethnicity"))

# Create overlay plot
glm_race_sex_plot <- ggplot(plot_data, aes(x = Year, color = race_ethnicity)) +
  geom_point(aes(y = observed_mean), size = 1.5, alpha = 0.6) +
  geom_line(aes(y = predicted_mean), size = 1.2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Observed and Predicted Mortality Rates by Race/Ethnicity (2000–2019)",
    x = "Year",
    y = "Mortality Rate (per 100,000)",
    color = "Race/Ethnicity"
  ) +
  scale_x_continuous(
    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
  ) +
  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.2, "cm"))

# Display
print(glm_race_sex_plot)

# Save plot (overwrite previous)
ggsave(
  here::here("results", "figures", "glm_race_sex_plot.png"),
  plot = glm_race_sex_plot,
  width = 7.5,
  height = 5.5,
  dpi = 300)

The GLM including race/ethnicity and sex as predictors highlights large disparities in predicted mortality rates. Black populations had the highest predicted rates, significantly higher than all other groups (e.g., β = 205.8, p \< 2e-16). Hispanic and Asian/Pacific Islander groups had lower mortality, though the Asian effect was not significant (p = 0.35). Males also showed significantly higher mortality than females (β = 20.3, p \< 2e-16). Notably, the “Overall” category behaves inconsistently in plots due to it being modeled as a separate group rather than a true average, likely affected by unmodeled age variation. This supports the need to model interactions and stratify further by age.

@@

invisible(saveRDS(glm_results_interaction, file = here::here("results", "output", "statistical-analysis", "glm_results_interaction.rds")))

# Plot the interaction model results
county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")
glm_interaction_plot <- ggplot(county_race_sex, aes(x = Year, y = predicted_mortalities_interaction, color = race_ethnicity)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "GLM Interaction Model: Mortality Trends by Year and Race", x = "Year", y = "Predicted Mortality Rate", color = "Race/Ethnicity") +
  scale_x_continuous(breaks = seq(min(county_race_sex$Year, na.rm = TRUE), max(county_race_sex$Year, na.rm = TRUE), by = 5)) +
  theme(legend.position = "right", legend.text = element_text(size = 11), legend.key.size = unit(0.4, "cm"), legend.spacing.y = unit(0.2, "cm"), axis.text.x = element_text(angle = 45, hjust = 1))

print(glm_interaction_plot)
invisible(ggsave(here::here("results", "figures", "glm_interaction_plot.png"), plot = glm_interaction_plot))

invisible(saveRDS(county_race_sex, file = here::here("results", "output", "statistical-analysis", "county_race_sex_with_preds.rds")))
```

```{r}
# Explore whether interactions exist between year and race/sex
glm_model_interaction <- glm(
  mortalities ~ Year * race_ethnicity + sex,
  data = county_race_sex,
  family = gaussian())

invisible(saveRDS(
  glm_results_interaction,
  file = here::here("results", "output", "statistical-analysis", "glm_results_interaction.rds")))

# Add interaction predictions
county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")

# Filter out "Overall" before plotting
plot_data <- county_race_sex %>%
  filter(race_ethnicity != "Overall")

# Create clean interaction plot
glm_interaction_plot <- ggplot(plot_data, aes(x = Year, y = predicted_mortalities_interaction, color = race_ethnicity)) +
  geom_line(size = 1.2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "GLM Interaction Model: Mortality Trends by Year and Race/Ethnicity",
    x = "Year",
    y = "Predicted Mortality Rate (per 100,000)",
    color = "Race/Ethnicity"
  ) +
  scale_x_continuous(
    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
  ) +
  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.2, "cm"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5))

# Display plot
print(glm_interaction_plot)

# Save cleaned image
ggsave(
  here::here("results", "figures", "glm_interaction_plot.png"),
  plot = glm_interaction_plot,
  width = 7.5,
  height = 5.5,
  dpi = 300)

# Save updated data
invisible(saveRDS(county_race_sex,file = here::here("results", "output", "statistical-analysis", "county_race_sex_with_preds.rds")))

```{r}
# Fit interaction model
glm_model_interaction <- glm(
  mortalities ~ Year * race_ethnicity + sex,
  data = county_race_sex,
  family = gaussian())

# Save model results
glm_results_interaction <- broom::tidy(glm_model_interaction)
print(glm_results_interaction)
invisible(saveRDS(
  glm_results_interaction,
  file = here::here("results", "output", "statistical-analysis", "glm_results_interaction.rds")))

# Add predicted values to dataset
county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")

# Filter out "Overall" before plotting
filtered_data <- county_race_sex %>%
  filter(race_ethnicity != "Overall")

# Summarize observed data by group and year
observed_summary <- filtered_data %>%
  group_by(Year, race_ethnicity) %>%
  summarise(observed_mean = mean(mortalities, na.rm = TRUE), .groups = "drop")

# Summarize predicted means
predicted_summary <- filtered_data %>%
  group_by(Year, race_ethnicity) %>%
  summarise(predicted_mean = mean(predicted_mortalities_interaction, na.rm = TRUE), .groups = "drop")

# Combine both summaries
plot_data <- left_join(observed_summary, predicted_summary, by = c("Year", "race_ethnicity"))

# Create overlay plot
glm_interaction_plot <- ggplot(plot_data, aes(x = Year, color = race_ethnicity)) +
  geom_point(aes(y = observed_mean), size = 1.5, alpha = 0.6) +  # Observed
  geom_line(aes(y = predicted_mean), size = 1.2) +               # Predicted
  theme_minimal(base_size = 12) +
  labs(
    title = "Observed and Predicted Mortality Rates by Race/Ethnicity (2000–2019)",
    x = "Year",
    y = "Mortality Rate (per 100,000)",
    color = "Race/Ethnicity"
  ) +
  scale_x_continuous(
    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
  ) +
  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.y = unit(0.2, "cm"))

# Display updated plot
print(glm_interaction_plot)

# Save image
ggsave(
  here::here("results", "figures", "glm_interaction_plot.png"),
  plot = glm_interaction_plot,
  width = 7.5,
  height = 5.5,
  dpi = 300)

# Save updated dataset
invisible(saveRDS(
  county_race_sex,
  file = here::here("results", "output", "statistical-analysis", "county_race_sex_with_preds.rds")))

Including an interaction between year and race revealed significantly divergent temporal trends. Mortality rates among Black individuals showed the steepest annual decline (β = –7.06, p \< 0.001), while American Indian and Alaska Native (AI/AN) populations exhibited the largest annual increase (β for year = 3.38, p \< 0.001, with positive main effect and significant negative interactions for other groups). Asian and Pacific Islander, Hispanic, and White populations also showed declining trends, though to a lesser extent. These results highlight substantial heterogeneity in how hypertension-related CVD mortality changed over time across racial/ethnic groups.

@@

# Save results
invisible(saveRDS(glm_age_group_results, file = here::here("results", "output", "statistical-analysis", "glm_age_group_results.rds")))

@@

# Save as table
invisible(saveRDS(model_fit_agegroup, here::here("results", "tables", "model_fit_agegroup.rds")))

@@

# Save plot
invisible(ggsave(here::here("results", "figures", "glm_age_plot.png"), 
       plot = glm_age_plot, width = 9, height = 6))

@@

This model identifies age-specific trends.Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends. Including an interaction between year and age group revealed that mortality rates are substantially higher in the 65+ age group across all years. While both age groups show increasing mortality over time, the slope is significantly steeper for older adults (interaction term β = 0.224, p \< 0.001), suggesting accelerated increases in mortality among those aged 65 and older.

@@

# Save as manuscript-ready table
invisible(saveRDS(slope_table_age_fmt, here::here("results", "tables", "slope_table_agegroup_fmt.rds")))

# Create and save polished PNG table for manuscript
slope_table_agegroup_gt <- slope_table_age_fmt %>%
  gt() %>%
  tab_header(
    title = "Estimated Annual Increase in Mortality by Age Group",
    subtitle = "Derived from GLM with Year × Age Group Interaction"
  ) %>%
  fmt_number(columns = c(Estimate, SE), decimals = 3) %>%
  cols_label(
    `Age Group` = "Age Group",
    Estimate = "Slope",
    SE = "Standard Error",
    CI = "95% CI"
  ) %>%
  opt_table_font(font = google_font("Roboto")) %>%
  tab_options(table.width = pct(80),             # shrink width
  table.font.size = "12",         # shrink font
  data_row.padding = px(2),          # reduce row height
  column_labels.font.weight = "bold")

gtsave(
  slope_table_agegroup_gt,
  filename = here::here("results", "tables", "slope_table_agegroup.png"))

slope_table_agegroup_gt

```{r}
# compare trends between groups (pairwise contrast)
slope_contrast <- contrast(em_trends_age, method = "pairwise")
summary(slope_contrast)

# Save
invisible(saveRDS(em_trends_age, file = here::here("results", "output", "statistical-analysis", "emtrends_age_group.rds")))
invisible(saveRDS(slope_contrast, file = here::here("results", "output", "statistical-analysis", "slope_contrast_age_group.rds")))

@@

# Save formatted contrast table
invisible(saveRDS(slope_contrast_age_fmt, here::here("results", "tables", "slope_contrast_agegroup_fmt.rds")))

@@

# Save figure to manuscript folder
invisible(ggsave(filename = here::here("results", "figures", "slope_plot_age_group.png"),
  plot = slope_plot_agegroup,
  width = 9, height = 6, dpi = 300))

# Save RDS for reproducibility
invisible(saveRDS(slope_plot_agegroup, file = here::here("results", "output", "statistical-analysis", "slope_plot_age_group.rds")))

@@

Estimated yearly changes in mortality rates were significantly higher for adults aged 65+ (slope = 1.26, 95% CI: 1.18–1.34) compared to those aged 35–64 (slope = 1.04, 95% CI: 0.96–1.12). The difference in slope between age groups was statistically significant (difference = -0.224, p \< 0.001), indicating a steeper rise in mortality among older adults.

@@

print(spatial_plot_all)
invisible(ggsave(here::here("results", "figures", "spatial_plot_all.png"), plot = spatial_plot_all))

@@

print(black_spatial_plot)
invisible(ggsave(here::here("results", "figures", "black_spatial_plot.png"), plot = black_spatial_plot))

@@

print(white_spatial_plot)
invisible(ggsave(here::here("results", "figures", "white_spatial_plot.png"), plot = white_spatial_plot))

@@

print(hispanic_spatial_plot)
invisible(ggsave(here::here("results", "figures", "hispanic_spatial_plot.png"), plot = hispanic_spatial_plot))

@@

print(native_spatial_plot)
invisible(ggsave(here::here("results", "figures", "native_spatial_plot.png"), plot = native_spatial_plot))

@@

print(asian_spatial_plot)
invisible(ggsave(here::here("results", "figures", "asian_spatial_plot.png"), plot = asian_spatial_plot))

```{r}
# Convert to sf object
county_race_sex_sf <- county_race_sex %>%
  st_as_sf(coords = c("X_long", "Y_lat"), crs = 4326, remove = FALSE)

# Create a reusable theme
spatial_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 13, face = "bold"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10)
  )

# 1. All races
spatial_plot_all <- ggplot() +
  geom_sf(data = county_race_sex_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of Mortality Rates – All Races", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
  spatial_theme
ggsave(here::here("results", "figures", "spatial_plot_all.png"), spatial_plot_all, width = 7, height = 5, dpi = 320)

spatial_plot_all

# 2. Black
black_race_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Black")
black_spatial_plot <- ggplot() +
  geom_sf(data = black_race_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of Mortality Rates – Black", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
  spatial_theme
ggsave(here::here("results", "figures", "black_spatial_plot.png"), black_spatial_plot, width = 7, height = 5, dpi = 320)

black_spatial_plot

# 3. White
white_race_sf <- county_race_sex_sf %>% filter(race_ethnicity == "White")
white_spatial_plot <- ggplot() +
  geom_sf(data = white_race_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of Mortality Rates – White", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
  spatial_theme
ggsave(here::here("results", "figures", "white_spatial_plot.png"), white_spatial_plot, width = 7, height = 5, dpi = 320)

white_spatial_plot

# 4. Hispanic
hispanic_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Hispanic")
hispanic_spatial_plot <- ggplot() +
  geom_sf(data = hispanic_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of Mortality Rates – Hispanic", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
  spatial_theme
ggsave(here::here("results", "figures", "hispanic_spatial_plot.png"), hispanic_spatial_plot, width = 7, height = 5, dpi = 320)

hispanic_spatial_plot

# 5. American Indian and Alaska Native
native_sf <- county_race_sex_sf %>% filter(race_ethnicity == "American Indian and Alaska Native")
native_spatial_plot <- ggplot() +
  geom_sf(data = native_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of Mortality Rates – American Indian and Alaska Native", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
  spatial_theme
ggsave(here::here("results", "figures", "native_spatial_plot.png"), native_spatial_plot, width = 7, height = 5, dpi = 320)

native_spatial_plot

# 6. Asian and Pacific Islander
asian_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Asian and Pacific Islander")
asian_spatial_plot <- ggplot() +
  geom_sf(data = asian_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Distribution of Mortality Rates – Asian and Pacific Islander", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
  spatial_theme
ggsave(here::here("results", "figures", "asian_spatial_plot.png"), asian_spatial_plot, width = 10, height = 5, dpi = 320)

asian_spatial_plot

@@

```{r}
# Explicit spacing and margins
combined_spatial_plot <- (
  spatial_plot_all + theme(plot.margin = margin(10, 10, 10, 10)) |
  black_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10))
) /
(
  white_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10)) |
  hispanic_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10))
) /
(
  native_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10)) |
  asian_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10))
)

ggsave(
  here::here("results", "figures", "combined_spatial_plot.png"),
  plot = combined_spatial_plot,
  width = 21,   # large print-friendly layout
  height = 15,
  dpi = 320
)

print(combined_spatial_plot)
```

These plots are best viewed as images outside of R due to size limits in R.

@@

### Evaluating spatial clustering of Mortality Rates.

### 2.3: Global Moran’s I – All Races Combined

@@

# Save to results/tables
invisible(saveRDS(moran_table, here::here("results", "tables", "global_morans_i_summary.rds")))
invisible(saveRDS(moran_result, file = moran_path))

@@

The Global Moran’s I statistic (I = 0.0435, p \< 0.001) indicates weak but statistically significant positive spatial autocorrelation in hypertension-related CVD mortality at the county level. There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S. counties — counties with high (or low) mortality tend to be near others with similarly high (or low) values. Although the magnitude is small (I = 0.0435), the pattern is considered statistically significant based on the p-value. This suggests that counties with high mortality rates tend to cluster geographically, rather than being randomly distributed. The presence of spatial autocorrelation supports further analysis using Local Moran’s I to identify specific high- and low-mortality clusters.

@@

# Save plot
invisible(ggsave(filename = here::here("results", "figures", "lisa_significance_map_centroids.png"),
  plot = lisa_point_plot,
  width = 10,
  height = 6))

@@

invisible(ggsave(here::here("results", "figures", "lisa_cluster_type_map.png"),
  plot = lisa_cluster_plot,
  width = 10, height = 6))

@@

#### Focusing at a State level Due to the size of the dataset it is challenging to work with the full dataset.

For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step. Our previous analysis support that there is a general distribution of cases here, even if there are not distinct hotspots.

@@

invisible(ggsave(here::here("results", "figures", "ga_lisa_cluster_plot.png"),
  plot = ga_lisa_plot,
  width = 8,
  height = 6))
```

@@

invisible(ggsave(filename = here::here("results", "figures", "ms_lisa_cluster_plot.png"),
  plot = ms_lisa_plot,
  width = 8,
  height = 6))

@@

# Save plot
invisible(ggsave(filename = here::here("results", "figures", "tx_lisa_cluster_plot.png"),
  plot = tx_lisa_plot,
  width = 9, height = 6))
```

@@

```{r}
# Paths for saving spatial weights
ky_knn_path <- here::here("results", "output", "ky_knn_neighbors.rds")
ky_lw_path  <- here::here("results", "output", "ky_spatial_weights.rds")

@@

# Save
invisible(ggsave(here::here("results", "figures", "lisa_cluster_kentucky.png"),
  plot = lisa_ky_plot, width = 9, height = 6))

@@

# Save the result
invisible(saveRDS(ky_moran_result, here::here("results", "output", "ky_global_morans_i.rds")))

@@

The Local Moran’s I analysis detected significant spatial clustering of mortality rates across U.S. counties. Nationally, over 40% of counties showed statistically significant clustering at p \< 0.05. However, most of these represented spatial outliers—counties with high (or low) mortality rates surrounded by contrasting neighbors—rather than concentrated hotspots. Only a small number of “High–High” clusters (hotspots) were identified, indicating that extreme spatial concentrations of high mortality are rare.

@@

#### 4.4.3 Comparing the LASSO and RF models to the Null model
